\bf Drift Removal for Time Series Data Using Quantile Trend Filtering

### 2019-01-12 Manuscript/detrendify_JCGS.tex
figure updates, new classification metrics

@@

	\subsection{Quantile Regression}

	The classic least squares regression is notoriously sensitive to outliers. One remedy to blunt the influence of outliers is to compute the least absolute deviations (LAD) solution in place of the least squares one. Given a design matrix $\mathbf{X} \in \mathbb{R}^{n \times p}$ and continuous responses $y \in \mathbb{R}^n$, we estimate a regression vector $\theta \in \mathbb{R}^p$ so that $\mathbf{X}\theta$ is a good approximation of $y$. The LAD estimator is a solution to the problem
	\begin{eqnarray}
		\underset{\theta}{\min}\; \frac{1}{n}\lVert y - \mathbf{X}\theta \rVert_1.
	\end{eqnarray}
	The above optimization problem generalizes the notion of the median of a collection of numbers. A median $\mu$ of $n$ reals $y_1, \ldots, y_n$ is the minimizer of the function
	\begin{eqnarray}
		f(u) & = & \frac{1}{n}\sum_{i=1}^n \lvert y_i - \theta \rvert.
	\end{eqnarray}
	Recall that the median is the 50th percentile or 0.5-quantile, namely half of the $y_i$ are less than or equal to $\mu$ and the other half is greater than or equal to $\mu$. The median can be generalized to arbitrary $\tau$-quantiles for $\tau \in (0,1)$ to give us quantile regression \citep{Koenker1978}.

	First define the so-called "check function"
	\begin{eqnarray}
		\rho_\tau(\Delta) & = & \begin{cases}
			\tau \Delta & \Delta \geq 0 \\
			-(1-\tau)\Delta & \Delta < 0 \\
		\end{cases}
	\end{eqnarray}
	Then the $\tau$th quantile of the $y_i$ is a minimizer of the function
	\begin{eqnarray}
		f_\tau(\theta) & = & \frac{1}{n}\sum_{i=1}^n \rho_\tau(y_i - \theta).
	\end{eqnarray}
	Returning to the regression context, we can generalize LAD regression to quantile regression, namely computing the minimizer of the function
	\begin{eqnarray}
		f_\tau(\theta) & = & \frac{1}{n}\sum_{i=1}^n \rho_\tau(y_i - \langle x_i \mid \theta \rangle),
	\end{eqnarray}
	where $x_i \in \mathbb{R}^p$ denotes the $i$th row of $\mathbf{X}$.

	%% ----------------------------------------------------------------------
	%% Trend Filtering
	%% ----------------------------------------------------------------------
	\subsection{Trend Filtering}

	%% ----------------------------------------------------------------------
	%% Quantile Trend Filtering
	%% ----------------------------------------------------------------------
	\section{Methods}

@@

	To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose an alternating direction method of multipliers (ADMM) algorithm for solving large problems in a piecewise fashion. The ADMM algorithm, described by \cite{boyd2011distributed}, relies on the idea of dual ascent. If we consider the equality optimization problem 
	\begin{align}
	&\text{minimize } f(x)\\
	&\text{subject to } Ax = b
	\end{align}
	with variable $x\in \mathbf{R}^n$, where $A \in \mathbf{R}^{mxn}$ and $f: \mathbf{R}^n \rightarrow \mathbf{R}$ is convex. The corresponding Lagrangian function is 
	\begin{equation}
	L(x,\omega) = f(x) + \omega^T(Ax-b)
	\end{equation}
	and the corresponding dual problem  is 
	\begin{equation}
	\text{maximize } \inf_x L(x,\omega)
	\end{equation}
	Assuming that strong duality holds (Slater's constraint: there exists a solution in the interior of the domain), the optimal values of the primal and dual variables are the same. The dual ascent method, uses gradient ascent to solve the dual problem and thus the primal problem using the iterates
	\begin{align}
	&x^{k+1} = \arg\min L(x, \omega^k)\\
	&\omega^{k+1} = \omega^k + \alpha^k(Ax^{k+1} - b)
	\end{align}
	where $\alpha^k > 0$ is a step size and $k$ is the iteration counter. In many cases (including our problem), the $x$ update fails because the problem is unbounded below. To remedy this issue the augmented Lagrangian was introduced:
	\begin{equation}
	L_{\gamma}(x,\omega) = f(x)+\omega^T(Ax-b) + \frac{\gamma}{2}||Ax-b||_2^2
	\end{equation}
	where $\gamma > 0$ is called the penalty parameter. Applying dual ascent to the modified problem results in the method of multipliers algorithm which converges under more general conditions:
	\begin{align}
	&x^{k+1} = \arg\min L_\gamma(x, \omega^k)\\
	&\omega^{k+1} = \omega^k + \gamma(Ax^{k+1} - b)
	\end{align}
	Consider a new optimization problem with a separable objective function. 
	\begin{align}
	&\text{minimize } f_1(x_1) + f_2(x_2)\\
	&\text{subject to } A_1x_1 + A_2x_2 = b
	\end{align}
	where $x = (x_1, x_2)$, $i \in \{1,2\}$, and  $x_i \in \mathbf{R}^{n_i}$. The dual ascent updates can be expressed as 
	\begin{align}
	&x_i^{k+1} = \arg\min L_i(x_i, \omega^k)\\
	&\omega^{k+1} = \omega^k + \alpha^k(Ax^{k+1} - b)
	\end{align}
	however we lose the separability if the augmented Lagrangian is used instead, i.e. the augmented Lagrangian can not be separated into functions of $x_i$.  The Alternating Direction Method of Multipliers (ADMM) addresses this problem and maintains separability while using the augmented Lagrangian to improve convergence through the following updates
	\begin{align}
	&x_1^{k+1} = \arg\min_{x_1} L_{\gamma}(x_1, x_2^k, \omega^k)\\
	&x_2^{k+1} = \arg\min_{x_2} L_{\gamma}(x_1^{k+1}, x_2, \omega^k)\\
	&\omega^{k+1} = \omega^k + \alpha^k(A_1x_1^{k+1} + A_2x_2^{k+1} - b)
	\end{align}

	We now consider the quantile regression trend filtering problem,  
	\begin{equation}
	\text{minimize } \rho_\tau(y-\theta) + \lambda||D^{(k)}\theta||_1\\
	\end{equation}
	where $y,\theta \in \mathbf{R}^n$, $y$ is the observed data, $\rho_{\tau}(u) = \sum_i(\tau - I(u_i < 0))u_i$ is the check loss function corresponding to quantile level $\tau$ and $D^{(k)}$ is the discrete differencing matrix of order $k$ and $\lambda$ is a tuning parameter that controls smoothness. We first re-parameterize $\phi = y-\theta$ so the problem is
	\begin{equation}
	\text{minimize } \rho_\tau(\phi) + \lambda||D^{(k)}(y-\phi)||_1\\
	\end{equation}
	We further divide $\phi$ order to solve smaller problems: Defining
	\begin{align}
	&\phi_1 = (\phi_{11}, \phi_{12})\\
	&\phi_2 = (\phi_{21}, \phi_{22}, \phi_{23})\\
	&\phi_3 = (\phi_{31}, \phi_{32})\\
	&\phi = (\phi_{11}, \phi_{12}=\phi_{21}, \phi_{22}, \phi_{23}=\phi_{31}, \phi_{32}) \\
	\end{align}
	Dividing $y$ similarly, the problem then becomes 
	\begin{align}
	&\text{minimize } \sum_{i=1}^3 \rho_\tau(\phi_i) + \lambda||D^{(k)}(y_i-\phi_i)||_1\\
	&\text{subject to: } \phi_{12}=\phi_{21}, ~~ \phi_{23}=\phi_{31}\\
	\end{align}
	We can further simplify by defining 
	\begin{align}
	&\overline{\phi} = (\phi_{11}, \frac{\phi_{12}+\phi_{21}}{2}, \phi_{22}, \frac{\phi_{23}+\phi_{31}}{2}, \phi_{32}) \\
	&\overline{\phi_1} = (\phi_{11}, \frac{\phi_{12}+\phi_{21}}{2})\\
	&\overline{\phi_2} = ( \frac{\phi_{12}+\phi_{21}}{2}, \phi_{22}, \frac{\phi_{23}+\phi_{31}}{2})\\
	&\overline{\phi_3} = (\frac{\phi_{23}+\phi_{31}}{2}, \phi_{32})
	\end{align}
	so the problem becomes
	\begin{align}
	&\text{minimize } \sum_{i=1}^3 \rho_\tau(\phi_i) + \lambda||D^{(k)}(y_i-\phi_i)||_1\\
	&\text{subject to: } \phi_{i}=\overline{\phi_i}\\
	\end{align}
	The augmented Lagrangian for this problem is 
	\begin{align}
	&L_\gamma(\phi_1,\phi_2, \phi_3, \overline{\phi_1}, \overline{\phi_2}, \overline{\phi_3}, \omega) = \\
	&\sum_{i=1}^3 \rho_\tau(\phi_i) + \lambda||D^{(k)}(y_i-\phi_i)||_1 + \omega_i^T(\phi_i - \overline{\phi_i}) + \frac{\gamma}{2}||\phi_i - \overline{\phi_i}||_2^2
	\end{align}
	The ADMM updates are then given by
	\begin{align}
	&\phi_i^{k+1} = \arg\min_{\phi_i}\rho_\tau(\phi_i) + \lambda||D^{(k)}(y_i-\phi_i)||_1 + \omega_i^{kT}(\phi_i - \overline{\phi_i}^k) + \frac{\gamma}{2}||\phi_i - \overline{\phi_i}^k||_2^2\\
	&\omega_i^{k+1} = \omega_i^{k} + \gamma(\phi_i^{k+1} - \overline{\phi_i}^{k+1})
	\end{align}
	The $\phi_i$ updates can be obtained using a quadratic program solver such as Gurobi and can be obtained in parallel.

@@

	\begin{figure}[!h] 
		\caption{Windows fit separately compared to simultaneous fit, signal present.}
		\includegraphics[width = \linewidth]{Figures/overlapping_windows2.png}
	\end{figure}

	\begin{figure}[!h] 
		\caption{Windows fit with 300 iterations of ADMM, signal present.}
		\includegraphics[width = \linewidth]{Figures/admm_windows2.png}
	\end{figure}

@@

	\begin{figure}[h!]
		\caption{RMSEs compared to the simulated baseline function.}
		\includegraphics[width = \linewidth]{Figures/peaks_mse.png}
	\end{figure}

@@

@@

	\input{application_missclass.tex}	
	\input{application_signal.tex}

>>>>

	A variety of approaches have been proposed for estimating quantile trends. \cite{Koenker1978} was the first to propose the use of the check loss function to estimate conditional quantiles. Given a design matrix $\mathbf{X} \in \mathbb{R}^{n \times p}$ and continuous responses $y \in \mathbb{R}^n$, we estimate a regression vector $\theta \in \mathbb{R}^p$ so that $\mathbf{X}\theta$ is a good approximation of the $\tau$th quantile of $y$ conditional on $x$. The regression vector $\theta$ is estimated by finding the minimizer of $\rho_{\tau}(y-X\theta)$ where $\rho_{\tau}(y) = \sum_{i=1}^n y_i(1-\mathbf{I}(y_i<0))$ is the check-loss function and $\mathbf{I}$ is the indicator function.

@@

	\section{Methods}

@@

	To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose an alternating direction method of multipliers (ADMM) algorithm for solving large problems in a piecewise fashion. The ADMM algorithm is fully described by \cite{boyd2011distributed}. We apply the consensus ADMM algorithm to the the quantile regression trend filtering problem, 
	\begin{equation}
	\text{minimize } \rho_\tau(y-\theta) + \lambda||D^{(k)}\theta||_1\\
	\end{equation}
	where $y,\theta \in \mathbf{R}^n$, $y$ is the observed data, $\rho_{\tau}(u) = \sum_i(\tau - I(u_i < 0))u_i$ is the check loss function corresponding to quantile level $\tau$ and $D^{(k)}$ is the discrete differencing matrix of order $k$ and $\lambda$ is a tuning parameter that controls smoothness. We first re-parameterize $\phi = y-\theta$ so the problem is
	\begin{equation}
	\text{minimize } \rho_\tau(\phi) + \lambda||D^{(k)}(y-\phi)||_1\\
	\end{equation}
	We further divide $\phi$ order to solve smaller problems: Defining
	\begin{align}
	&\phi_1 = (\phi_{11}, \phi_{12})\\
	&\phi_2 = (\phi_{21}, \phi_{22}, \phi_{23})\\
	&\phi_3 = (\phi_{31}, \phi_{32})\\
	&\phi = (\phi_{11}, \phi_{12}=\phi_{21}, \phi_{22}, \phi_{23}=\phi_{31}, \phi_{32}) \\
	\end{align}
	Dividing $y$ similarly, the problem then becomes 
	\begin{align}
	&\text{minimize } \sum_{i=1}^3 \rho_\tau(\phi_i) + \lambda||D^{(k)}(y_i-\phi_i)||_1\\
	&\text{subject to: } \phi_{12}=\phi_{21}, ~~ \phi_{23}=\phi_{31}\\
	\end{align}
	We can further simplify by defining 
	\begin{align}
	&\overline{\phi} = (\phi_{11}, \frac{\phi_{12}+\phi_{21}}{2}, \phi_{22}, \frac{\phi_{23}+\phi_{31}}{2}, \phi_{32}) \\
	&\overline{\phi_1} = (\phi_{11}, \frac{\phi_{12}+\phi_{21}}{2})\\
	&\overline{\phi_2} = ( \frac{\phi_{12}+\phi_{21}}{2}, \phi_{22}, \frac{\phi_{23}+\phi_{31}}{2})\\
	&\overline{\phi_3} = (\frac{\phi_{23}+\phi_{31}}{2}, \phi_{32})
	\end{align}
	so the problem becomes
	\begin{align}
	&\text{minimize } \sum_{i=1}^3 \rho_\tau(\phi_i) + \lambda||D^{(k)}(y_i-\phi_i)||_1\\
	&\text{subject to: } \phi_{i}=\overline{\phi_i}\\
	\end{align}
	The augmented Lagrangian for this problem is 
	\begin{align}
	&L_\gamma(\phi_1,\phi_2, \phi_3, \overline{\phi_1}, \overline{\phi_2}, \overline{\phi_3}, \omega) = \\
	&\sum_{i=1}^3 \rho_\tau(\phi_i) + \lambda||D^{(k)}(y_i-\phi_i)||_1 + \omega_i^T(\phi_i - \overline{\phi_i}) + \frac{\gamma}{2}||\phi_i - \overline{\phi_i}||_2^2
	\end{align}
	The ADMM updates are then given by
	\begin{align}
	&\phi_i^{k+1} = \arg\min_{\phi_i}\rho_\tau(\phi_i) + \lambda||D^{(k)}(y_i-\phi_i)||_1 + \omega_i^{kT}(\phi_i - \overline{\phi_i}^k) + \frac{\gamma}{2}||\phi_i - \overline{\phi_i}^k||_2^2\\
	&\omega_i^{k+1} = \omega_i^{k} + \gamma(\phi_i^{k+1} - \overline{\phi_i}^{k+1})
	\end{align}
	The $\phi_i$ updates can be obtained using a quadratic program solver such as Gurobi and can be obtained in parallel.

@@

	We conducted timing experiments to show the advantages of using our ADMM algorithm on large datasets. We use 2 windows for each data size with an overlap of 2000 and two iterations of the algorithm after the initial estimates.

	\begin{figure}[!h] 
		\caption{Timing experiments comparing quantile trend filtering on a single window versus 2 windows by data size.}
		\includegraphics[width = \linewidth]{Figures/Fig_timing_experiment.png}
	\end{figure}

@@

@@

	\begin{figure}[h!]
		\caption{Class averaged accuracy by threshold, data size, and method (1 is best 0.5 is worst).}
		\includegraphics[width = \linewidth]{Figures/peaks_CAA.png}
	\end{figure}

	\begin{figure}[h!]
		\caption{F1 score by threshold, data size, and method (1 is best 0 is worst).}
		\includegraphics[width = \linewidth]{Figures/peaks_F1.png}
	\end{figure}

	\begin{figure}[h!]
		\caption{Precision by threshold, data size, and method (true positive over true positives + false positives).}
		\includegraphics[width = \linewidth]{Figures/peaks_precision.png}
	\end{figure}

	\begin{figure}[h!]
		\caption{Recall by threshold, data size, and method (true positive over true positives + false negatives).}
		\includegraphics[width = \linewidth]{Figures/peaks_recall.png}
	\end{figure}

@@

	\input{application_metrics.tex}

	\includegraphics[width = \linewidth]{Figures/corrected_rugplot.png}
	\includegraphics[width = \linewidth]{Figures/signal_plot.png}
